---
title: "Week 4: Data Management"
subtitle:  "Session 2"

header-includes:
  \usepackage{multicol}
  
output:
  beamer_presentation:
    
    # highlight: tango
#     theme: "default"
#     colortheme: "spruce"
#     fonttheme: "serif"
#     slide_level: 2
#     incremental: false
# urlcolor: blue
# classoption: t
---

```{r setup, echo=FALSE, results='hide', message=FALSE}
library(knitr)
require(ggplot2)
require(data.table)


iris_dt = data.table(iris)


# # opts_knit$set(root.dir = "G:/Dropbox/C/Courses class/NRC290b/NRC_290b_2018/")
# #opts_knit$set(root.dir = "C:/Users/csutherland/Dropbox/C/Courses class/NRC290b/NRC_290b_2018")
# 
# opts_chunk$set(prompt = FALSE, comment = NA, mysize=TRUE, size="\\scriptsize")
# hook1 <- function(x){ gsub("```\n*```r*\n*", "", x) }
# hook2 <- function(x){ gsub("```\n+```\n", "", x) }
# knit_hooks$set(document = hook2)
# knitr::knit_hooks$set(mysize = function(before, options, envir) {
#   if (before) 
#     return(options$size)
# })
```


# iClicker questions



\begin{multicols}{2}
\null \vfill
\vfill \null
\columnbreak
\includegraphics[width = 0.35\textwidth]{../slide_images/iClicker_logo.png}
\end{multicols}





# Distributions

Your Gardener book mentions the Normal distribution in chapter 4, but there is no prior discussion about what *distribution* means!

What is a distribution?

We're going to build an intuitive concept of a distribution from the ground up.




# Distributions

[example of variability with a real dataset]

```{r}
iris_dt[, sep_wid_jit := jitter(Sepal.Width)]
iris_dt[, sep_len_jit := jitter(Sepal.Length)]
iris_dt[, x_jit := jitter(rep(0, nrow(iris_dt)))]
iris_dt

```



```{r}
ggplot(iris_dt, aes(x = sep_wid_jit, y = x_jit)) + geom_point() + ylim(-0.1, 0.1) +
  ylab("") + theme(axis.ticks.y = element_blank(), axis.text.y = element_blank())





plot(
  y = jitter(rep(0, nrow(iris))),
  x = jitter(iris$Sepal.Length),
  ylim = c(-0.1, 0.1),
  
  main = "Iris sepal widths")





```







# The Normal Distributions





